---
title: "Experiment 007: Bacterial Trajectory Following Systemic Infection with *P. rettgeri* "
author: "by Barbara Black"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
      code_folding: hide
      number_sections: true
      toc: true
      toc_float: true
      toc_depth: 3
      fig_caption: true
      theme: united
      css: style.css
      highlight: zenburn
editor_options: 
  chunk_output_type: console
---

<style>

/* Change font size of the title */
h1.title {
  font-size: 22px;
}

/* Define the frames */
div.blue { background-color:#e6f0ff; border-radius: 1px; padding: 10px; }
div.orange { background-color:#fff8dc; border-radius: 5px; padding: 20px; margin: 10px 0;}
div.green { background-color:#00bbc6; border-radius: 5px; padding: 20px;
  border: 1px solid #000; margin: 10px 0;}

/* Headers */ 
h1 { font-size: 18px; font-weight: bold;}
h2 { font-size: 16px; font-weight: bold; margin: 10px 0;}
h3 { font-size: 14px; font-weight: bold; margin: 10px 0;}

/* Change font size of the table of contents */
TOC { font-size: 12px; }
</style>

```{r setup, include=FALSE}
## set global options
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, results = 'hide', dpi = 300)
```

```{r git-setup}
# # Initialise a local Git repo
# usethis::use_git()
# 
# # Connect local Git repo to GitHub
# usethis::use_github()
```

# Background 

# Code {.tabset}

## Functions

```{r}
### LOAD FUNCTIONS
### Function to change code chunk size in output
#   Note: Automatically applied when chunk runs
def.chunk.hook = knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x = def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

### Function to read contents from listed files
#   Note: Modified from read_and_standardise() function used in survival datasets
read_contents = function(file) {
  df = read_csv(file.path(path, file))
  return(df)
}

### Function to check and correct miscounted deaths
#   Note: Use with grouped dataset and within apply()
correct_status = function(row) {
  status = row['status']
  n = row['n']
  censored = row['censored']
  status = as.numeric(status)
  n = as.numeric(n)
  censored = as.numeric(censored)
  if (status > n) {
    return(n - censored)
  } else {
    return(status)
  }
}

### Function to extract table from mixed effects model
extract_coxme_table = function (mod){
    beta = fixef(mod)
    nvar = length(beta)
    nfrail = nrow(mod$var) - nvar
    se = sqrt(diag(mod$var)[nfrail + 1:nvar])
    z = round(beta/se, 2)
    p = format(as.numeric(pchisq((beta/se)^2, 1,lower.tail = FALSE)), 4)
    table = data.frame(cbind(beta,se,z,p))
    return(table)
}

# Function to extract survival data to plot
ggplotprep2 <- function(x, times){
  d <- data.frame(condition=rep(names(x$strata), x$strata), time=x$time, survival=x$surv, upper=x$upper, lower=x$lower)
  fillup0 <- function(s) rbind(c(condition=s, time=0, survival=1, upper=1, lower=1), d[d$condition==s, ], deparse.level = 0)
  
  indexes <- function(x, time) {
    if(x%in%time) return(x)
    return(floor(time[which.min(abs(time[time<x]-x))]))
  }
  
  fillup <- function(s) {
    d.temp <- d[d$condition==s, ]
    time <- as.numeric(d.temp$time)
    id <- sapply(times, indexes, time=time)
    d.temp <- d.temp[match(id, time), ]
    d.temp$time <- times
    return(d.temp)
  }
  
  if(times[1]==0) d <- do.call("rbind", sapply(names(x$strata), fillup0, simplify=F))
  d <- do.call("rbind", sapply(names(x$strata), fillup, simplify=F))
  clean.name <- function(name) unlist(lapply(strsplit(as.character(name), split="="), function(x) x[2]))
  d <- data.frame(Condition=clean.name(d$condition), Time=as.numeric(d$time), Survival=as.numeric(d$survival), upper=as.numeric(d$upper), lower=as.numeric(d$lower))
  return(d)
}
```

## Libraries

```{r}
### LOAD LIBRARIES
library(openxlsx)
library(survival)
library(ggsurvfit)
library(car)
library(gtsummary)
library(survminer)
library(janitor)
library(ggsci)
library(ggtext)
library(mltools)
library(broom)
library(coxme)
library(formatR)
library(tidyverse)
library(cowplot)
```

## Data Import and Processing

```{r}
### IMPORT DATA
#   Set path and read data files
path = here::here("data/input")

data_files = list.files(path, 
                        pattern = "*.csv")

#   Standardise all columns as character variables across all data files
data_list = data.frame(filename = data_files) %>% 
  mutate(file_contents = map(filename, 
                             ~ read_contents(.)))

#   Unnest the data list into a single data frame
data_raw = unnest(data_list, 
                  cols = file_contents)

#   Print the result
print(data_raw)
```

```{r}
### MODIFY COUNTS DATA
data = data_raw %>% 
  separate(`sample_ID;BLUD_or_SPBL;WELL;Volume;Dilution;BACTERIA;NA`, 
           into = c("sample_ID", "BLUD_or_SPBL", "WELL", "Volume", "Dilution", "BACTERIA", "NA"),
           sep = ";") 

data = data %>% 
  separate(`BLUD_or_SPBL`, into = paste0("word", 1:7), sep = "_") %>% 
  select(!c(filename, word2, word3, word4)) %>% 
  rename("BLUD_or_SPBL" = "word1",
         "Time_post_injection" = "word5",
         "DF" = "word6",
         "Plate" = "word7") %>% 
  mutate(Plate = paste0("Plate", sub("^0+", "", Plate)),
         Antibiotic = "None") %>% 
  mutate(sample_PlateID = paste(sample_ID, Plate, sep = "_")) %>% 
  select(-sample_ID) %>% 
  select(sample_PlateID, BLUD_or_SPBL, Plate, Antibiotic, everything()) %>% 
  relocate(Time_post_injection, .after = last_col()) %>% 
  mutate(Plate = sub("Plate", "", Plate)) 

data_rep = data %>% 
  group_by(Plate, DF, Time_post_injection) %>% 
  summarise(Total = n()) %>% 
  select(-Total) %>% 
  group_by(DF, Time_post_injection) %>%
  mutate(Rep = row_number())
  
data_join = data %>% 
  left_join(data_rep, by = c("Plate", "DF", "Time_post_injection")) 

count_data = data_join %>% 
  select(!c(DF, Time_post_injection)) %>% 
  relocate(Rep, .after = Antibiotic)

### EXPORT COUNTS DATA
write.csv(count_data, "data/output/COUNTS_config_007_prett_modified.csv",                       
          row.names = FALSE)
```

```{r}
### MODIFY PLAN DATA
plan_data = data_join %>% 
  rename(sample_ID = sample_PlateID) %>% 
  mutate(Sex = if_else(grepl("^[A-D]", WELL), "Female", "Male")) %>% 
  mutate(Age = if_else(grepl("(1|2|5|6|9|10)$", WELL), "Young", "Old")) %>% 
  separate(WELL, into = c("Row", "Col"), sep = "(?<=\\D)(?=\\d)") %>% 
  select(sample_ID, Sex, Time_post_injection, Plate, Row, Col, Age) %>% 
  mutate(Col = as.numeric(Col)) %>% 
  mutate(Time_post_injection_modified = case_when(
    grepl("t1-t3", Time_post_injection) & Col %in% c(1, 2, 3, 4) ~ 1,
    grepl("t1-t3", Time_post_injection) & Col %in% c(5, 6, 7, 8) ~ 2,
    grepl("t1-t3", Time_post_injection) & Col %in% c(9, 10, 11, 12) ~ 3,
    
    grepl("t4-t6", Time_post_injection) & Col %in% c(1, 2, 3, 4) ~ 4,
    grepl("t4-t6", Time_post_injection) & Col %in% c(5, 6, 7, 8) ~ 5,
    grepl("t4-t6", Time_post_injection) & Col %in% c(9, 10, 11, 12) ~ 6,
    
    grepl("t7-t8", Time_post_injection) & Col %in% c(1, 2, 3, 4) ~ 7,
    grepl("t7-t8", Time_post_injection) & Col %in% c(5, 6, 7, 8) ~ 8,
    
    grepl("t9-t11", Time_post_injection) & Col %in% c(1, 2, 3, 4) ~ 9,
    grepl("t9-t11", Time_post_injection) & Col %in% c(5, 6, 7, 8) ~ 10,
    grepl("t9-t11", Time_post_injection) & Col %in% c(9, 10, 11, 12) ~ 11,
    
    grepl("t12-t13", Time_post_injection) & Col %in% c(1, 2, 3, 4) ~ 12,
    grepl("t12-t13", Time_post_injection) & Col %in% c(5, 6, 7, 8) ~ 13,
    
    grepl("t14-t16", Time_post_injection) & Col %in% c(1, 2, 3, 4) ~ 14,
    grepl("t14-t16", Time_post_injection) & Col %in% c(5, 6, 7, 8) ~ 15,
    grepl("t14-t16", Time_post_injection) & Col %in% c(9, 10, 11, 12) ~ 16,
    
    grepl("t24-t48", Time_post_injection) & Col %in% c(1, 2, 3, 4) ~ 24,
    grepl("t24-t48", Time_post_injection) & Col %in% c(5, 6, 7) ~ 48,
    grepl("t24-t48", Time_post_injection) & Col %in% c(8, 9, 10, 11) ~ 24,
    grepl("t24-t48", Time_post_injection) & Col == 12 ~ 48)) %>% 
  
  mutate(Sex = case_when(Time_post_injection_modified == 48 & Row == "E" ~ "Female",
                         TRUE ~ Sex)) %>% 
  select(-Time_post_injection) %>% 
  rename(Time_post_injection = Time_post_injection_modified)

### EXPORT PLAN DATA
write.csv(plan_data, "data/output/PLAN_config_007_prett_modified.csv",                       
          row.names = FALSE)
```

```{r}
### CHECK count_data$BACTERIA FOR NAs, ZEROES AND EMPTY ROWS 
count_na_bacteria = which(is.na(count_data$BACTERIA))
na_rows_bacteria = count_data[count_na_bacteria, ]  # Note: There are no rows with NAs.
na_rows_bacteria



count_zero_bacteria = which(count_data$BACTERIA == "0")
zero_rows_bacteria = count_data[count_zero_bacteria, ] # Follow up: Do all rows where BACTERIA == 0 correspond to rows where NA == 1, and vice-versa?
zero_rows_bacteria



count_empty_bacteria = which(count_data$BACTERIA == "")
empty_rows_bacteria = count_data[count_empty_bacteria, ] # Follow up: Do all rows where BACTERIA == "" correspond to rows where NA == "", and vice-versa?
empty_rows_bacteria


### CHECK count_data$`NA` FOR NAs, ZEROES AND EMPTY ROWS 
count_na_NA = which(is.na(count_data$`NA`))
na_rows_NA = count_data[count_na_NA, ] # Note: There are no rows with NAs.
na_rows_NA



count_zero_NA = which(count_data$`NA` == "0")
zero_rows_NA = count_data[count_zero_NA, ] # Follow up: Do all rows where BACTERIA > 0 correspond to rows where NA == "0", and vice-versa?
zero_rows_NA



count_empty_NA = which(count_data$`NA` == "")
empty_rows_NA = count_data[count_empty_NA, ] # Same question already asked.
empty_rows_NA



### FOLLOW UP QUESTIONS AND ANSWERS
# Question: Do all rows where BACTERIA == 0 correspond to rows where NA == 1? (Answer: Yes)
all(count_data$`NA`[which(count_data$BACTERIA == "0")] == 1)

# Question: Do all rows where NA == 1 correspond to rows where BACTERIA == 0? (Answer: Yes)
all(count_data$BACTERIA[which(count_data$`NA` == 1)] == "0")



# Question: Do all rows where NA == "" correspond to rows where BACTERIA == ""? (Answer: Yes)
all(count_data$BACTERIA[which(count_data$`NA` == "")] == "")

# Question: Do all rows where BACTERIA == "" correspond to rows where NA == ""? (Answer: Yes)
all(count_data$`NA`[which(count_data$BACTERIA == "")] == "")



# Question: Do all rows where BACTERIA > 0 correspond to rows where NA == "0", and vice-versa? (Answer: Yes)
all(as.numeric(count_data$`NA`)[which(as.numeric(count_data$BACTERIA) > 0)] == 0)

# Question: Do all rows where NA > 0 correspond to rows where BACTERIA == "0", and vice-versa? (Answer: Yes)
all(as.numeric(count_data$BACTERIA)[which(as.numeric(count_data$`NA`) > 0)] == 0)



### CHECK PLATE NUMBERS FOR WHEN THE VALUES IN BOTH BACTERIA AND NA ARE EMPTY
empty_bacteria_na = count_data %>% 
  filter(BACTERIA == "" & `NA` == "") 

plate_number_empty = empty_bacteria_na %>% 
  group_by(Plate) %>% 
  summarise()                       # Conclusion: When both BACTERIA == "" and NA == "" in a given row, that means the same as BACTERIA == "0" and NA == "1";
                                    #             in other words, no samples were added to these wells.
                                    #             This is only seen across A-G rows because when selecting "transfer to column" in CountDrops, 
                                    #             the NA was originally marked in the H well (i.e., NA == "1").



### CHECK PLATE NUMBERS AND REPLICATES
count_data %>% 
  group_by(Plate, Rep) %>% 
  summarise()                       # Note: Plates 79 and 80 (associated with T24 and T48) only show one replicate in the dataset.
                                    #       That is because both replicates were done in the same plate (see plate map for details).
                                    #       Both plates are related to the same time points, but each one has a different dilution factor.

```

# Bacterial Trajectory

```{r, fig.dim = c(4,4), fig.align='center', out.width="60%"}
my_data = count_data %>% 
  left_join(plan_data, by = join_by("sample_PlateID" == "sample_ID")) %>% 
  select(sample_PlateID, Rep, Age, Sex, Dilution, Volume, BACTERIA, Time_post_injection) %>% 
  mutate(across(c(Dilution, Volume, BACTERIA, Time_post_injection), as.numeric),
         across(where(is.character), as.factor),
         cfu_per_ml = (BACTERIA * Dilution * 1000)/(Volume * 250),
         log2_cfu = log2(cfu_per_ml),
         log2_cfu = ifelse(is.infinite(log2_cfu), NA, log2_cfu))

my_data %>% 
  filter(Time_post_injection < 20) %>% 
  ggplot(aes(Time_post_injection, log2_cfu, colour = Age)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  #facet_grid(~Sex) +
  labs(x = "Time (hours post-infection)",
       y = expression(log[2] ~ "(CFU/mL)")) +
  scale_x_continuous(limits = c(0, 16),
                     breaks = c(seq(0, 16, by = 2))) +
  scale_y_continuous(limits = c(0, 20)) +
  theme_cowplot() +
  theme(aspect.ratio = 1) 
```

